---
title: "Hg Fish_HF"
author: "CBG"
date: "12 11 2021"
output: html_document
---
```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)
```
### Packages
The following packages are required
- dplyr (for summarising data)
```{r load required packages, include=FALSE}
Packages <- c("plyr", "dplyr", "graphics", "ggplot2", "GGally", "bestglm", "magrittr", "mgcv", "tidymv", "MuMIn", "tidyr")
lapply(Packages, library, character.only = TRUE)
```

```{r VIF function for GAM}
samvif <- function(mod){
  # mod is an mgcv object
  # this function calculates the variance inflation factors for GAM as no one else has written code to do it properly
  # this is used to summarise how well the GAM performed
  mod.sum <- summary(mod)
  s2 <- mod$sig2 # estimate of standard deviation of residuals
  X <- mod$model # data used to fit the model
  n <- nrow(X) # how many observations were used in fitting?
  v <- -1 # omit the intercept term, it can't inflate variance
  varbeta <- mod.sum$p.table[v,2]^2 # variance in estimates
  varXj <- apply(X=X[,row.names(mod.sum$p.table)[v]],MARGIN=2, var) # variance of all the explanatory variables
  VIF <- varbeta/(s2/(n-1)*1/varXj) # the variance inflation factor, obtained by rearranging
  # var(beta_j) = s^2/(n-1) * 1/var(X_j) * VIF_j
  
  VIF.df <- data.frame(variable=names(VIF),
                       vif=VIF, 
                       row.names=NULL)
  return(VIF.df)
}
```

```{r Function for testing linearity of distributions}
#Function MyMultipanel.ggp2: Mixed effects models and extensions in ecology with R. (2009).
#Zuur, AF, Ieno, EN, Walker, N, Saveliev, AA, and Smith, GM. Springer.
MyMultipanel.ggp2 <- function(Z, varx, vary, 
                              ylab = "Response variable",
                              addSmoother = FALSE,
                              addRegressionLine = FALSE,
                              addHorizontalLine = FALSE) {
  K <- length(varx)
  MyData <- data.frame(Y = rep(as.vector(as.matrix(Z[,vary])), K),
                       X = as.vector(as.matrix(Z[, varx])),
                       Var = rep(varx, each = nrow(Z))) 
  library(ggplot2)
  p <- ggplot(MyData, aes(y = Y, x = X))
  p <- p + geom_point() + ylab(ylab) + xlab("Covariates")
  p <- p + theme(text = element_text(size=15))
  if (addSmoother == TRUE) {
    p <- p + geom_smooth(se = TRUE, col = "black", lwd = 1)
  }
  if (addRegressionLine == TRUE) {
    p <- p + geom_smooth(se = TRUE, col = "black", lwd = 1, method = "lm")
  }
  if (addRegressionLine == TRUE) {
    p <- p + geom_smooth(se = TRUE, col = "black", lwd = 1, method = "lm")
  }
  if (addHorizontalLine == TRUE) {
    p <- p + geom_hline(yintercept = 0)
  }
  p <- p + facet_wrap(~ Var, scales = "free_x")
  suppressMessages(print(p)) 	
}

#To make plots individually to individual optimization
#Function for å definere antall siffer i akse
fmt_dcimals <- function(decimals=1){
  function(x) format(x,nsmall = decimals,scientific = FALSE)
}
```
## Model Hg in Fish (Aim)
Look at potential change in fish Hg at three different (Mjøsa, Ellasjøen, Femunden,Tollreien, and Breidtjern) over time (1990-2020) with the following parameters as potential explanatory variables: 
- fish descriptors: length, weight, sex, age, stable C and N isotopes
- annual average air temperature and precipitation
- annual average Hg-deposition and possibly Hg-concentration in air
- TOC data available for Breidtjern and Pt for Vorma (Mjøsa/Femunden)

## Procedure
1) look at the THg data, should all years be included (influential points)?
2) look at the variables: 
  - influential points? 
  - Collinearity? if so, those parameters should not be included together in the model. 
  - Linear distributions? - if looking no, check distribution of residuals, if no -> use GAM (to avoid data transformation)
  - Temporal autocorrelation?

### Generalized additive model (GAM)
- use models better suited to the situation, rather than torturing the data to fit a simplified modelling scheme
- incorporates nonlinear forms of the predictors
- using a smooth function of the predictor variables, which can take on a great many form
- allows for nonlinear relationships between the covariates and the target variable y
- GAMs are suspectible to extrapolation, as is every statistical method ever created
- being additive and not normally including interactions between covariates
The sorts of smooths we fit in mgcv are (typically) penalized smooths; we choose to use some number of basis functions k, which sets an upper limit on the complexity — wiggliness — of the smooth, and then we estimate parameters for the model by maximizing a penalized log-likelihood. The log-likelihood of the model is a measure of the fit (or lack there of), while the penalty helps us avoid fitting overly complex smooths.
```{r Load data}
THg <- read.table("THg_AllFishData.txt", sep = "\t", header = TRUE)
Vari <- read.table("All_Variables.txt", sep = "\t", header = TRUE)
TOC <- read.table("Breidtjern_TOC.txt", sep = "\t", header = TRUE)
Pt <- read.table("Mjosa_Femunden_Pt_Vorma.txt", sep = "\t", header = TRUE)
```
### Data preparation
- convert some columns from character to numeric
- annual average of THg in fish data
- merge fish data with other variables
- THg in fish, Ellasjøenx1000 higher than other sites
- make new Hg_dep column with lag of 1 year
```{r Data preparation}
THg$Length <- as.numeric(THg$Length)
Vari$Temp <- as.numeric(Vari$Temp)

THg_X = THg %>%
  group_by(Year, Lake) %>%
  select(Length:d13C, Age) %>% # select variables to summarise, from and to
  dplyr::summarise(across(everything(), .f = list(mean = mean), na.rm = TRUE)) 

Vari_X = Vari %>%
  group_by(Year, Site) %>%
  select(Hg_dep:Temp) %>% # select variables to summarise, from and to
  dplyr::summarise(across(everything(), .f = list(mean = mean), na.rm = TRUE))

TOC_X = TOC %>%
  group_by(Year, Lake) %>%
  select(Value) %>% # select variables to summarise, from and to
  dplyr::summarise(across(everything(), .f = list(mean = mean), na.rm = TRUE))

Pt_X = Pt %>%
  group_by(Year) %>%
  select(Concentration_TOC) %>% # select variables to summarise, from and to
  dplyr::summarise(across(everything(), .f = list(mean = mean), na.rm = TRUE))

THg_X$Site = THg_X$Lake
Vari_X$Site <- revalue(Vari_X$Site, c("Mj?sa" = "Mjøsa", "Ellasj?en"="Ellasjøen"))
THg_X$Site <- revalue(THg_X$Site, c("Mj?sa" = "Mjøsa", "Ellasjoen"="Ellasjøen"))
THg_all <- merge(THg_X, Vari_X, by=c("Year", "Site"), all.x=TRUE)
THg_all <- THg_all[,(-3)]

THg_all$Hg_dep_mean2[2:nrow(THg_all)] <- THg_all$Hg_dep_mean[1:(nrow(THg_all) - 1)]
```

```{r Subset data based on site, add TOC/Pt data where applicable}
Brei <- subset(THg_all, Site == 'Breidtjern')
Brei$Lake = "Brei"
BreiPt <- merge(Brei, Pt_X, by="Year")
Ella <- subset(THg_all, Site == 'Ellasjøen')
Ella$Lake = "Ella"
Fem <- subset(THg_all, Site == 'Femunden')
Fem$Lake = "Fem"
Mjo <- subset(THg_all, Site == 'Mjøsa')
Mjo$Lake = "Mjo"
MjoTOC <- merge(Mjo, TOC_X, by="Year", all.x=TRUE)
MjoTOC$TOC = MjoTOC$Value_mean
Toll <- subset(THg_all, Site == 'Tollreien')
Toll$Lake = "Toll"
```
### Explore data
- boxplot of data per year per site (look for extreme observations)
  -Mjøsa early observations (<1990) should be excluded
- correlation plot of variables from each site (look for collinearity)
- For several of the sites there is colinearity between several of the variables, e.g. weight and length. Correlating variables should not be included in the same model. 

```{r Make boxplot per lake, without and with observations}
THg2<-THg[!(THg$Lake=="Randsfjorden"),]
THg2$Lake <- revalue(THg2$Lake, c("Mj?sa" = "Mjøsa", "Ellasjoen"="Ellasjøen"))

ggplot(THg2, aes(x=Year, y=THg, fill=Lake, group=Year))+
  geom_boxplot()+
  facet_grid(rows = vars(Lake), scales="free")+
  scale_y_continuous(labels = fmt_dcimals(1))+
  labs(title = "", y ="THg (mg/kg)")

# https://www.r-graph-gallery.com/89-box-and-scatter-plot-with-ggplot2.html
ggplot(THg2, aes(x=Lake, y=log10(THg), fill=Lake)) +
    geom_boxplot() +
    geom_jitter(color="black", size=0.4, alpha=0.9)+
    scale_y_continuous(limits = c(-1.7, 0.7), labels = fmt_dcimals(1))+
    labs(title = "", y ="log(THg (mg/kg))")+
  geom_abline(aes(colour="black"),intercept=log10(0.02), slope=0)+
  geom_abline(aes(colour="blue", linetype=d),intercept=log10(0.5), slope=0)
  #scale_colour_manual(name='My Lines', values=c("blue"))
```
```{r explore data: Correlation analysis - plotting data per site}
BreiX <- subset(BreiPt, select=c(Length_mean, Weight_mean, Hg_dep_mean, Hg_dep_mean2, Prec_mean, Concentration_TOC_mean))
#ggpairs(BreiX,title="Breidtjern")
EllaX <- subset(Ella, select=c(Length_mean, Weight_mean, Hg_dep_mean, Hg_dep_mean2, Prec_mean, Temp_mean))
#ggpairs(EllaX,title="Ellasjøen")
FemX <- subset(Fem, select=c(Length_mean, Weight_mean, d15N_mean, d13C_mean, Hg_dep_mean, Hg_dep_mean2, Prec_mean, Temp_mean))
#ggpairs(FemX,title="Femunden")
MjoX <- subset(MjoTOC, select=c(Length_mean, Weight_mean, d15N_mean, Prec_mean, Temp_mean, Hg_dep_mean, Hg_dep_mean2, TOC))
#ggpairs(MjoX,title="Mjøsa")
TollX <- subset(Toll, select=c(Length_mean, Weight_mean, Hg_dep_mean, Hg_dep_mean2, Prec_mean, Temp_mean))
#ggpairs(TollX,title="Tollreien")
```
### General data adaptations, checking diagnostics
http://www.sthda.com/english/articles/39-regression-model-diagnostics/161-linear-regression-assumptions-and-diagnostics-in-r-essentials/
- qq plot for residuals, remover "outliers"
- transformation of the target variable, in case of heteroscedacity (not because of non-normality!) some might take the log
- it usually a good idea to scale predictor variables so that the effect is more meaningful

## Case Study: Mjøsa
**Extreme observations:** exclude observations before 1990 since they are few, likely uncertain, but will have weight (?)
**Collinearity:** between length and weight, and with d15N, and between Hg-dep and d15N (strong!)
            Confirmed using VIF-test of the model. Scores > 10, parameter should be excluded from model
Linear distribution: 
Temporal autocorrelation: results show p only slightly higher than 0.05

General for model selection:
- method "REML"
- basis: tp or cr (?)
- start by including all terms smooth. The k value could be set as low as 3. Remember that a smooth adds 10 parameters which could cause problem for datasets with limited observations (degres of freedom...)
- look at the results and if effective degrees of freedom =1, you know that term can be un-smooth
- If response not normally distributed. To avoid transformation use gamma distribution
- gam.check: if p-value is low and edf close to k' <- number of basis function is ok
- k never higher than number of datapoints
- maximum number of basis functions limits the maximum amount of non-linearity
```{r Mjøsa - Looking at data an possible collinearity}
ggpairs(MjoX,title="Mjøsa")
MjoX1 <- subset(MjoTOC, select=c(Year, Length_mean, Weight_mean, THg_mean,d15N_mean, Prec_mean, Temp_mean, Hg_dep_mean,Hg_dep_mean2, TOC))
MjoX_long <- gather(MjoX1, Variable, measurement, Length_mean:TOC, factor_key=TRUE)

#plotting the variables separately
ggplot(MjoX_long, aes(y = measurement, x = Year))+ 
  geom_point(size = 3)+
  facet_wrap(Variable~., scales="free")+
  labs(title ="Mjøsa Variable data",x = "Year", y= "")

Mjoo <- subset(MjoTOC, Year>1990)
MyVar  <- c("Year", "Length_mean", "Weight_mean","d15N_mean", "Prec_mean", "Temp_mean", "Hg_dep_mean", "Hg_dep_mean2","TOC")
MyMultipanel.ggp2(Mjoo, 
                  varx = MyVar, 
                  vary = "THg_mean",
                  ylab = "Mercury",
                  addSmoother = TRUE,
                  addRegressionLine = FALSE,
                  addHorizontalLine = FALSE)

#For non-normal distributions, test residuals
#Non-linearity suggested for Year and d15N -> check with residuals
#Test1 <- uGamm(THg_mean~s(Year), random=list(Length_mean=~1), data=Mjoo)
#plot(Test1$gam,residuals=TRUE,all.terms=F, pages=1, pch=21, main = "Hg vs year")
#  abline(h=0.0, col="red") #non-linear
#gam.check(Test1$gam) #not normally distributed

#Test2 <- uGamm(THg_mean~s(Length_mean), random=list(Length_mean=~1),data=Mjoo)
#plot(Test2$gam,residuals=TRUE,all.terms=F, pages=1, pch=21, main = "Hg vs Weight")
#abline(h=0, col="red")
#gam.check(Test2$gam)
#Test3 <- uGamm(THg_mean~s(d15N_mean), random=list(Length_mean=~1),data=Mjoo)
#plot(Test3$gam,residuals=TRUE,all.terms=F, pages=1, pch=21, main = "Hg vs d15")
#abline(h=0, col="red")

#Test4 <- uGamm(THg_mean~s(Hg_dep_mean), random=list(Length_mean=~1),data=Mjoo)
#plot(Test4$gam,residuals=TRUE,all.terms=F, pages=1, pch=21, main = "Hg vs Hg dep")
#abline(h=0, col="red") #linear
```


```{r case study: Mjøsa}
#HGData$YEAR.c <- (HGData$YEAR - min(HGData$YEAR)+1) #centering year to get reasonable intercept
MjoX1 <- subset(MjoTOC, select=c(Year, Length_mean, Weight_mean, THg_mean,d15N_mean, Prec_mean, Temp_mean, Hg_dep_mean,Hg_dep_mean2, TOC))
MjoX_long <- gather(MjoX1, Variable, measurement, Length_mean:TOC, factor_key=TRUE)
#plotting the variables separately
ggplot(MjoX_long, aes(y = measurement, x = Year))+ 
  geom_point(size = 3)+
  facet_wrap(Variable~., scales="free")+
  labs(title ="Mjøsa Variable data",x = "Year", y= "")

Mjoo <- subset(MjoTOC, Year>1990)
#for ggpairs, need to select without Year
#ggpairs(MjooX,title="Mjøsa > 1990")

# E) Temporal Autocorrelation
#H0: model does not show lack of fit = "model is fine". p<0.05 rejects the H0. 
MjooX <- subset(Mjoo, select=c(Length_mean, Weight_mean, d15N_mean, Prec_mean, Temp_mean, Hg_dep_mean, Hg_dep_mean2,TOC))
acf(MjoX1$THg_mean) 
Box.test(MjoX1$THg_mean, lag = 1, type = c("Box-Pierce", "Ljung-Box"), fitdf = 0)

#centering year to get reasonable intercept
Mjoo$Year.c <- (Mjoo$Year - min(Mjoo$Year)+1)

#Scaling covariates for comparability
Mjoo$Weight_mean.sc <- scale(Mjoo$Weight_mean)
Mjoo$Hg_dep_mean.sc <- scale(Mjoo$Hg_dep_mean)
Mjoo$Hg_dep_mean2.sc <- scale(Mjoo$Hg_dep_mean2)
Mjoo$TOC.sc <- scale(Mjoo$TOC)
Mjoo$Temp_mean.sc <- scale(Mjoo$Temp_mean)
Mjoo$d15N_mean.sc <- scale(Mjoo$d15N_mean)
Mjoo$Prec_mean.sc <- scale(Mjoo$Prec_mean)

#Checking memory alocated to R
memory.size() ### Checking your memory size
memory.limit()
memory.limit(size=56000) 

# GAM model selection

#GLM use link function to model non-normal data
# Gamma versus Gaussian. Gamma lener seg mot den ene siden. 
# Om gamma og identity link: https://stats.stackexchange.com/questions/471374/generalized-linear-model-and-identity-link-whats-its-benefit
Hg.gam<-list()
Hg.gam[[1]] <- gam(THg_mean~ s(Year.c, k=3)+s(Weight_mean.sc, k=3)+s(Hg_dep_mean.sc, k=3)+s(Prec_mean.sc, k=3)+s(TOC.sc, k=3)+s(Temp_mean.sc, k=3)+s(d15N_mean.sc, k=3), 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))
Hg.gam[[2]] <- gam(THg_mean~ s(Year.c, k=4)+Weight_mean.sc+Hg_dep_mean.sc+s(Prec_mean.sc, k=4)+s(TOC.sc, k=4)+Temp_mean.sc+d15N_mean.sc, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))
Hg.gam[[3]] <- gam(THg_mean~ s(Year.c)+Weight_mean.sc+Hg_dep_mean.sc+Prec_mean.sc+TOC.sc+Temp_mean.sc+d15N_mean.sc, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))
Hg.gam[[4]] <- gam(THg_mean~ s(Year.c)+Weight_mean.sc+Hg_dep_mean2.sc+Prec_mean.sc+TOC.sc+Temp_mean.sc+d15N_mean.sc, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))
Hg.gam[[5]] <- gam(THg_mean~ s(Year.c)+Weight_mean.sc+Prec_mean.sc+TOC.sc+Temp_mean.sc+d15N_mean.sc, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))
Hg.gam[[6]] <- gam(THg_mean~ s(Year.c)+Weight_mean.sc+Hg_dep_mean2.sc+Prec_mean.sc+Temp_mean.sc+d15N_mean.sc, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))
Hg.gam[[7]] <- gam(THg_mean~ s(Year.c, k=8)+Weight_mean.sc+Hg_dep_mean2.sc+Prec_mean.sc+d15N_mean.sc, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))
Hg.gam[[8]] <- gam(THg_mean~ s(Year.c)+Weight_mean.sc+Hg_dep_mean2.sc+Prec_mean.sc, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))

plot(Hg.gam[[1]], rug = F, shade = T, main = "Visualise the need for s()", ylab = "")

final <- gam(THg_mean~ s(Year.c, k=8)+Weight_mean.sc+Hg_dep_mean2.sc+Prec_mean.sc+d15N_mean.sc, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))

finalX <- gam(THg_mean~ s(Year, k=8)+Weight_mean+Hg_dep_mean2+Prec_mean+d15N_mean, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))
finalY <- gam(THg_mean~ s(Year, k=8)+Weight_mean+Prec_mean+d15N_mean, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))

#+d15N_mean
model_pred = predict.gam(final,type="response",se.fit=T)
  df <- data.frame(matrix(unlist(model_pred)),stringsAsFactors=FALSE)
df$Type = c("pred", "pred", "pred", "pred", "pred", "pred", "pred", "pred", "pred", "pred", "pred", "pred", "pred", "pred", "pred", "pred", "SE", "SE", "SE", "SE", "SE","SE", "SE", "SE", "SE", "SE", "SE", "SE", "SE", "SE", "SE", "SE") 
df$Year = c(2000, 2001, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2000, 2001, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019) 
colnames(df) <- c("Value", "Type", "Year")
df2 <- spread(df, Type, Value)
head(df2)

df2$predgam = predict(finalX)

ggplot(Mjoo, aes(x=Year, y=THg_mean))+
  geom_point()+
  geom_line(data=df2, aes(x=Year, y=pred))
  geom_ribbon(data=df2, aes(ymin=pred-SE, ymax=pred+SE), alpha=0.1)

visreg(finalX, "Year.c")
  
library(tidyverse)

Hg.ga<-list()
Hg.ga[[1]] <- gam(THg_mean~ s(Year, k=3)+s(Weight_mean, k=3)+s(Hg_dep_mean, k=3)+s(Prec_mean, k=3)+s(TOC.sc, k=3)+s(Temp_mean, k=3)+s(d15N_mean, k=3), 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))
Hg.ga[[2]] <- gam(THg_mean~ s(Year, k=4)+Weight_mean+Hg_dep_mean+s(Prec_mean, k=4)+s(TOC, k=4)+Temp_mean+d15N_mean, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))
Hg.ga[[3]] <- gam(THg_mean~ s(Year)+Weight_mean+Hg_dep_mean+Prec_mean+TOC+Temp_mean+d15N_mean, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))
Hg.ga[[4]] <- gam(THg_mean~ s(Year)+Weight_mean+Hg_dep_mean2+Prec_mean+TOC+Temp_mean+d15N_mean, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))
Hg.ga[[5]] <- gam(THg_mean~ s(Year)+Weight_mean+Prec_mean+TOC+Temp_mean+d15N_mean, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))
Hg.ga[[6]] <- gam(THg_mean~ s(Year)+Weight_mean+Hg_dep_mean2+Prec_mean+Temp_mean+d15N_mean, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))
Hg.ga[[7]] <- gam(THg_mean~ s(Year, k=8)+Weight_mean+Hg_dep_mean2+Prec_mean+d15N_mean, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))
Hg.ga[[8]] <- gam(THg_mean~ s(Year)+Weight_mean+Hg_dep_mean2+Prec_mean, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))


model.sel(Hg.ga,rank = AIC ) 
gam.check(finalX, rep=1000)
summary(finalX)
samvif(Hg.gam[[7]])

  #predict_gam(finalX, values = list(f1 = c(0.5, 1, 1.5))) %>% #need library for predict_gam
  #ggplot(aes(x2, fit)) +
  #geom_smooth_ci(f1)
  

model.sel(Hg.gam,rank = AIC ) 
gam.check(Hg.gam[[7]], rep=1000)
summary(Hg.gam[[7]])
samvif(Hg.gam[[7]])

plot(Hg.gam[[7]],pages=1,residuals=TRUE,all.terms=TRUE,shade=TRUE,shade.col=2)
plot(Hg.gam[[7]],pages=1,seWithMean=TRUE) ## better coverage intervals
termplot(Hg.gam[[7]],terms="x0",se=TRUE)

model_p <- predict_gam(Hg.gam[[7]])
model_p

#when removed dep, model changed a lot. Removed TOC and kept dep - ok. 
print(Hg.gam[[7]])
#plot.gam(Hg.gam[[7]]) #plot uten observasjoner

library(tidymv)
library(voxel) #visualise GAM using ggplot
library(mgcViz)
b <- getViz(Hg.gam[[7]]) #convert GAM model to plot
o <- plot( sm(b, 1) )
o + l_fitLine(colour = "red") + l_rug(mapping = aes(x=x, y=y), alpha = 0.8) +
    l_ciLine(mul = 5, colour = "blue", linetype = 2) + 
    l_points(shape = 19, size = 5, alpha = 0.1) + theme_classic()+
  ggtitle("Mjøsa")
Hg.gam[[7]]

uff <- predict(Hg.gam[[7]], type="response", se.fit=TRUE)

plot(Mjoo$THg_mean,fitted(Hg.gam[[7]]))
```

```{r finding the link function}
M1 <- gam(THg_mean~ s(Year.c)++Weight_mean.sc+Hg_dep_mean2.sc+Prec_mean.sc+d15N_mean.sc, 
           select=TRUE, data=Mjoo, method="REML", family=Gamma(link="identity"))
M2 <- gam(THg_mean~ s(Year.c)++Weight_mean.sc+Hg_dep_mean2.sc+Prec_mean.sc+d15N_mean.sc, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="inverse"))
M3 <- gam(THg_mean~ s(Year.c)++Weight_mean.sc+Hg_dep_mean2.sc+Prec_mean.sc+d15N_mean.sc, 
             select=TRUE, data=Mjoo, method="REML", family=Gamma(link="log"))
M4 <- gam(THg_mean~ s(Year.c)++Weight_mean.sc+Hg_dep_mean2.sc+Prec_mean.sc+d15N_mean.sc, 
             select=TRUE, data=Mjoo, method="REML")
AIC(M1, M2, M3, M4)
```

- if everything is fine from the gam.check, the k-index would be 1 or greater. if too low we need to increase the ka value in the model. Residual not accounted for. 
- p-value if significantly different from the random test. should not be different (p  0.05)

##Case Study: Femunden
- the two observations around year 2000 are higher than the remaining, without data inbetween, and could be influential
- collinearity: length~weight, d13C_mean~d15N_mean, and weight~d13C
```{r case study Femunden}
#looking at collinearity
ggpairs(FemX,title="Femunden")
#looking at distribution of variables
MyVar  <- c("Year", "Length_mean", "Weight_mean","d15N_mean", "d13C_mean","Prec_mean", "Temp_mean", "Hg_dep_mean2")
MyMultipanel.ggp2(Fem, 
                  varx = MyVar, 
                  vary = "THg_mean",
                  ylab = "Mercury",
                  addSmoother = TRUE,
                  addRegressionLine = FALSE,
                  addHorizontalLine = FALSE)
#looking at the actual data of the variables
FemX1 <- subset(Fem, select=c(Year, Length_mean, Weight_mean, THg_mean,d15N_mean, d13C_mean, Prec_mean, Temp_mean, Hg_dep_mean2))
FemX_long <- gather(FemX1, Variable, measurement, Length_mean:Hg_dep_mean2, factor_key=TRUE)
#plotting the variables separately
ggplot(FemX_long, aes(y = measurement, x = Year))+ 
  geom_point(size = 3)+
  facet_wrap(Variable~., scales="free")+
  labs(title ="Mjøsa Variable data",x = "Year", y= "")

# E) Temporal Autocorrelation
#H0: model does not show lack of fit = "model is fine". p<0.05 rejects the H0. 
acf(FemX1$THg_mean) 
Box.test(FemX1$THg_mean, lag = 1, type = c("Box-Pierce", "Ljung-Box"), fitdf = 0)

#Centering and standardizing before mdoel fit
#centering year to get reasonable intercept
#Mjoo$Year.c <- (Mjoo$Year - min(Mjoo$Year)+1)

#Scaling covariates for comparability
#Mjoo$Weight_mean.sc <- scale(Mjoo$Weight_mean)
#Mjoo$Hg_dep_mean.sc <- scale(Mjoo$Hg_dep_mean)
#Mjoo$Hg_dep_mean2.sc <- scale(Mjoo$Hg_dep_mean2)
#Mjoo$TOC.sc <- scale(Mjoo$TOC)
#Mjoo$Temp_mean.sc <- scale(Mjoo$Temp_mean)
#Mjoo$d15N_mean.sc <- scale(Mjoo$d15N_mean)
#Mjoo$Prec_mean.sc <- scale(Mjoo$Prec_mean)

#fra VIF-test utgår C13. Valgte vekt for å ha likt som Mjøsa.
#Få observasjoner så kun Year som får s(). Kan testes senere?
#Fra summary: Temp dårligst p-value
nrow(Fem)
Hg.g<-list()
Hg.g[[1]] <- gam(THg_mean~ s(Year, k=4)+Weight_mean+Hg_dep_mean2+Prec_mean+d15N_mean+Temp_mean,  select=TRUE, data=Fem, method="REML", family=Gamma(link="identity"))
Hg.g[[2]] <- gam(THg_mean~ s(Year, k=4)+Weight_mean+Hg_dep_mean2+Prec_mean+d15N_mean,  select=TRUE, data=Fem, method="REML", family=Gamma(link="identity"))
Hg.g[[3]] <- gam(THg_mean~ s(Year, k=4)+Weight_mean+Prec_mean+d15N_mean,  select=TRUE, data=Fem, method="REML", family=Gamma(link="identity"))
Hg.g[[4]] <- gam(THg_mean~ s(Year, k=4)+Weight_mean+d15N_mean,  select=TRUE, data=Fem, method="REML", family=Gamma(link="identity"))
Hg.g[[5]] <- gam(THg_mean~ Year+Weight_mean+d15N_mean,  select=TRUE, data=Fem, method="REML", family=Gamma(link="identity"))
Hg.g[[6]] <- gam(THg_mean~ Weight_mean+d15N_mean,  select=TRUE, data=Fem, method="REML", family=Gamma(link="identity"))




model.sel(Hg.g,rank = AIC ) 
gam.check(Hg.g[[5]], rep=1000)
summary(Hg.g[[6]])
samvif(Hg.g[[2]])

visreg(Hg.g[[4]], "Year")

plot(Hg.gam[[7]],pages=1,residuals=TRUE,all.terms=TRUE,shade=TRUE,shade.col=2)
plot(Hg.gam[[7]],pages=1,seWithMean=TRUE) ## better coverage intervals
termplot(Hg.gam[[7]],terms="x0",se=TRUE)
```





##Case Study: Ellasjøen
- no apparent strict extreme observations
- weight and length significantly strongly correlated, should not be in the same model. All others ok. 
- weight seems better than length at explaining (from the model), but both good
- temporal autocorrelation seems ok
- including all as s() with low k, seems like temp is not significant, trying without the s(). Not significant without either.
- edf for s(Year) much higher than for the other. Reason to skip s() for the other? Try and then compare the models. 
- AIC model 1 slightly better than 2. Model 3 much better than 2 and 1. 
- Conclusion: 

### Uncertainties in the procedure (compared to the polarbear study)
- test for multicollinearity 
- test for interaction between terms
- checking linearity of residuals of variables (plot not working)
- what is tested by  ID.NR? some randamization?
- here, not tested generalized mnixed effect models